---
title: "ST558: Module2 R Activity"
author: "Philip Ourso"
date: "October 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(gridExtra)
library(corrplot)
```

# Create vectors in R using c()  
  
1. Create the vector [1  4  9  12] using the ‘c()’ function, and name it ‘newVec1’.  
    The vector is created and displayed below.    
    
```{r}
newVec1 = c(1, 4, 9, 12)
newVec1
```

2. Create another object:  
    > newVec2 <- (1:4)^2  
    Have R check whether newVec1 and newVec2 are identical.  
    
    The vector is created and compared below.    
    
```{r}
newVec2 = c(1:4)^2
all(newVec2 == newVec1)
```

# Create matrices in R using cbind(), rbind(), and matrix()  

3. Use the ‘cbind()’ command to create the following matrix and name it ‘newMat1’:  
    
    The matrix is created below.  
```{r}
newMat1 = cbind(
  c(1, 4, 2),
  c(0, 7, -3),
  c(5, 2, 1)
)
newMat1
```

4. Use the ‘rbind()’ command to create the following matrix and name it ‘newMat2’:  
    
    The matrix is created below.  
```{r}
newMat2 = rbind(
  c(1, 1, 2),
  c(4, 7, 2),
  c(5, -3, -1)
)
newMat2
```

5. Compare ‘newMat1’ and ‘newMat2’ to identify which positions are not equal.  
  
    The element-wise comparison is below.  
  
```{r}
newMat1 == newMat2
```

6. Use the ‘matrix()’ command to create the following matrix and name it ‘newMat3’.  
  
    The matrix is created below.  
    
```{r}
newMat3 = matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, ncol = 2)
newMat3
```

# Produce the transpose of a matrix using the t() command

7. Compare ‘newMat1’ to the transpose of ‘newMat2’ to identify which positions are not equal.  
    
    The element-wise comparison is shown below.  
    
```{r}    
newMat1 == t(newMat2)
```

8. How could you use R to check whether a matrix is symmetric? Give code for checking whether ‘newMat2’ is symmetric.    
    
    A symmetric matrix is equal to its transposition, so a comparison of all elements of a matrix to its transposition would indicate symmetry.  
    
```{r}
all(newMat2 == t(newMat2))
```

# Extract the diagonal elements using diag().

9. Create a diagonal matrix named ‘diagMat1’ with diagonal entries equal to the diagonal entries of ‘newMat1’.  
    
    The diagonal matrix is created below.  
    
```{r}
diagMat1 = diag(c(diag(newMat1)))
diagMat1
```

10. Create the (5 x 5) identity matrix and name it ‘ident5’.  
    
    The identity matrix is created below.  
    
```{r}
ident5 = diag(5)
ident5
```

# Perform matrix addition  
    
11. Calculate the sum of ‘newMat1’ and ‘newMat2’.  
    
    The sum is below.  
    
```{r}
newMat1 + newMat2
```

# Multiply a matrix by a constant  

11. Multiply ‘newMat3’ by 7.  
    
    The product is below.   
   
```{r}
newMat3 * 7
```

# Perform matrix multiplication

12. Create the (3 x 3) identity matrix I3, and name it ‘ident3’.  Perform matrix multiplication to find the product I3(newMat3).   
    
    The product is below.  
    
```{r}
ident3 = diag(3)
ident3 %*% newMat3
```

13. What happens when you try to perform matrix multiplication to find the product newMat3(I3)? Why is that?  
    
    When the product is attempted, R indicates that the arguments have non-conformable dimensions. newMat3 has 2 columns, but ident3 has 3 rows, and hence matrix multiplication can't be performed.    
    
```{r}
dim(newMat3)
dim(ident3)
#newMat3 %*% ident3

```

# Calculate the inverse of a matrix  

14. Find the matrix inverse of ‘newMat1’.  Name this matrix ‘newMat1.inv’ and report its value.  
    
    The inverse of the matrix is below.  
    
```{r}
newMat1.inv = solve(newMat1)
newMat1.inv
```
    
15. Demonstrate that the matrix product of ‘newMat1’ and ‘newMat1.inv’ is the (3 x 3) identity matrix, no matter which order you perform the multiplication.  
    
    Both products produce the identity matrix.  
    
```{r}
round(newMat1 %*% newMat1.inv)

round(newMat1.inv %*% newMat1)
```

# Calculate the determinant and trace of a matrix.  

16. What happens if you try to find the determinant of a matrix that is not square? Demonstrate an example.  
    
    The below example produces an error in which it is noted that the argument matrix is not square.   
   
```{r}
newMat9 = cbind(c(1,2,3), c(4,5,6))
#det(newMat9)
```

17. What happens if you try to find the trace of a matrix that is not square? Demonstrate an example.  
    
    In this case, no error message is produced, and a row is dropped to make a square matrix, with a value returned for the trace.  
    
```{r}
newMat9
diag(newMat9)
sum(diag(newMat9))
```
    
18. Calculate the determinant of ‘newMat2’.  
    
    The determinant is produced by the below code.  

```{r}
det(newMat2)
```

19. Calculate the trace of ‘newMat2’.  
    
    The trace is produced by the below code.  

```{r}
sum(diag(newMat2))
```

# Calculate the eigendecomposition of a square matrix.  

20. Create the following matrix and name it ‘newMat2’
Find the eigendecomposition of this matrix.  What is the eigenvector corresponding to the smallest eigenvalue?  
    
    The below code produces the eigenvector corresponding to the smallest eigenvalue via the which.min() function.   
   
```{r}
newMat2 = cbind(c(6.8,2.4,0), c(2.4,8.2,0), c(0,0,0.5))
newMat2.eig = eigen(newMat2)
#newMat2.eig$values
#newMat2.eig$vectors
newMat2.eig$vectors[,which.min(newMat2.eig$values)]
```

## Appendix: R code
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```
